---
title: "Session 1"
subtitle: "Getting to Know Your Data"
institute: "Global TIES for Children"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      countIncrementalSlides: false
      highlightLines: true
      highlightStyle: solarized-light
---
class: middle, center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(xaringanthemer)
style_mono_accent(
  text_font_google = google_font("Montserrat", "300", "300i"),
  code_font_google = google_font("Fira Mono")
)
library(magrittr)
library(flair)
```

# Loading your data

## ðŸ—„âž¡ ðŸ’»

---

## Understanding R syntax

R is built around **functions**, bits of atomic code that accept inputs and return an output.

--

A function looks like this:

```{r, eval=FALSE}
function_name(argument_1, argument_2, parameter = TRUE)
```

--

Functions can be bundled together and shared using **packages**. To access a function from a specific package, we use the `::` operator.

```{r, eval=FALSE}
package.name::function_name(argument_1, argument_2, parameter = TRUE)
```

---

## Understanding R syntax

Often you frequently reuse functions from a particular package, and using `::` can get tedious and ugly to read. R allows you to **attach** those packages using the `library()` function so you can just refer to the function name.

```{r, eval=FALSE}
library(package.name)

# Refers to `function_name` in `package.name`
function_name(argument_1, argument_2, parameter = TRUE)
```

---

# Tools for loading data

There are many file formats out there, but these are tools we recommend for file types that we commonly work with:

```{r, eval=FALSE}
readr::read_csv()
haven::read_dta()
openxlsx::read.xlsx()
```

**readr** and **haven** are both part of the "_tidyverse_", a suite of packages that have a consistent, easy-to-use, and easy-to-understand design.

**openxlsx** is a standalone package that is a successor to the **xlsx** package, an older utility that has a dreaded _Java dependency_. Use xlsx at your own peril!

---

# Let's load our data!

Michael has graciously provided an anonymized example intervention dataset for us to use.

It is located in the `data` directory (folder) at the root of this project. I'll use the **here** package to create paths that are _relative to the project root_.

--

I'll attach the packages we need for now:

```{r}
library(readr)
library(here)
```

Note how **here** tells us where it thinks the project root is! It recursively searches upward for a directory with a `DESCRIPTION` or `.Rproj` file. 

**here** is very handy utility for reproducibility since it _localizes_ paths to the project -- and not expanding to the full path on _your_ computer.

---

# Let's load our data!

Time to load the data:

```{r}
dat <- read_csv(here("data/sample_intervention_dataset.csv"))
```

---

## Two things to note

.pull-left[

R uses the `<-` operator to _assign_ computed values to symbols (variables) to be used later on. 

```{r, eval=FALSE}
x <- exp(-(y - mu)^2)
```

In our data's case, the results of reading the CSV file are assigned to the `dat` symbol.

]

.pull-right[

`readr::read_csv` prints out a _column specification_ so you can understand how it interpreted the data. This is handy to diagnose some import problems (usually related to dates).

```{r, echo=FALSE}
dat <- read_csv(here("data/sample_intervention_dataset.csv"))
```

]

---

# Looking at our data

After loading our data, it's useful to high-level view of its content to make sure that it loaded correctly.

We'll use the `str()` and `print()` functions here, both part of base R.

---

# Looking at our data

```{r}
str(dat)
```

---

# Looking at our data

The _class_ of this `dat` object is a `data.frame`. `data.frame`s are R's representation of tabular data. `dat` is also a `tbl_df` ("tibble"), which is the tidyverse's version of a `data.frame` that is a little bit more stringent (i.e. safe) than base R's version, and it has a prettier `print()` method:

```{r}
print(dat)
```

---

# Data types in R

You may have noticed a bunch of three-letter abbreviations next to the variable names in `print(dat)` and the variable content in `str(dat)`. These are the _type signatures_ for each variable:

- Double (`dbl` / `num`): All real numbers, like $50$, $23.193$, or $1.24\times10^{-8}$
- Character (`chr`): Text/strings, written with single- or double-quotes (e.g. `"a string"`)

There are a couple more that you will encounter regularly:

- Integer (`int`): Integers and counting numbers. No decimals and scientific notation here!
- Logical (`lgl`): Boolean data, written as `TRUE` or `FALSE`.
- Factor (`fct`): Factors, R's representation of categorical data. They look like character data with `print()`, but under the hood they are integers. These should only exist in cleaned, ready-for-analysis data!

---

# Recap

---
class: middle, center

# Working with your data

## ðŸ—„ðŸ”
