---
title: "Session 1"
subtitle: "Getting to Know Your Data"
institute: "Global TIES for Children"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      countIncrementalSlides: false
      highlightLines: true
      highlightStyle: solarized-light
---
class: middle, center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(xaringanthemer)
style_mono_accent(
  text_font_google = google_font("Montserrat", "300", "300i"),
  code_font_google = google_font("Fira Mono")
)
library(magrittr)
library(flair)
```

<style>
:root {
  --font12: 12pt;
  --font10: 10pt;
}

.font12 > table {
  font-size: var(--font12);
}

.font12 > p {
  font-size: var(--font12);
}

.font12 > pre .remark-code-line {
  font-size: var(--font12);
}

.font10 > table {
  font-size: var(--font10);
}

.font10 > p {
  font-size: var(--font10);
}

.font10 > pre .remark-code-line {
  font-size: var(--font10);
}
</style>

# Loading your data

## ðŸ—„âž¡ ðŸ’»

---

## Understanding R syntax

R is built around **functions**, bits of atomic code that accept inputs and return an output.

--

A function looks like this:

```{r, eval=FALSE}
function_name(argument_1, argument_2, parameter = TRUE)
```

--

Functions can be bundled together and shared using **packages**. To access a function from a specific package, we use the `::` operator.

```{r, eval=FALSE}
package.name::function_name(argument_1, argument_2, parameter = TRUE)
```

---

## Understanding R syntax

Often you frequently reuse functions from a particular package, and using `::` can get tedious and ugly to read. R allows you to **attach** those packages using the `library()` function so you can just refer to the function name.

```{r, eval=FALSE}
library(package.name)

# Refers to `function_name` in `package.name`
function_name(argument_1, argument_2, parameter = TRUE)
```

---

# Tools for loading data

There are many file formats out there, but these are tools we recommend for file types that we commonly work with:

```{r, eval=FALSE}
readr::read_csv()
haven::read_dta()
openxlsx::read.xlsx()
```

**readr** and **haven** are both part of the "_tidyverse_", a suite of packages that have a consistent, easy-to-use, and easy-to-understand design.

**openxlsx** is a standalone package that is a successor to the **xlsx** package, an older utility that has a dreaded _Java dependency_. Use xlsx at your own peril!

---

# Let's load our data!

Michael has graciously provided an anonymized example intervention dataset for us to use.

It is located in the `data` directory (folder) at the root of this project. I'll use the **here** package to create paths that are _relative to the project root_.

--

I'll attach the packages we need for now:

```{r}
library(readr)
library(here)
```

Note how **here** tells us where it thinks the project root is! It recursively searches upward for a directory with a `DESCRIPTION` or `.Rproj` file. 

**here** is very handy utility for reproducibility since it _localizes_ paths to the project -- and not expanding to the full path on _your_ computer.

---

# Let's load our data!

Time to load the data:

```{r}
dat <- read_csv(here("data/sample_intervention_dataset.csv"))
```

---

## Two things to note

.pull-left[

R uses the `<-` operator to _assign_ computed values to symbols (variables) to be used later on. 

```{r, eval=FALSE}
x <- exp(-(y - mu)^2)
```

In our data's case, the results of reading the CSV file are assigned to the `dat` symbol.

]

.pull-right[

`readr::read_csv` prints out a _column specification_ so you can understand how it interpreted the data. This is handy to diagnose some import problems (usually related to dates).

```{r, echo=FALSE}
dat <- read_csv(here("data/sample_intervention_dataset.csv"))
```

]

---

# Looking at our data

After loading our data, it's useful to high-level view of its content to make sure that it loaded correctly.

We'll use the `str()` and `print()` functions here, both part of base R.

---

# Looking at our data

```{r}
str(dat)
```

---

# Looking at our data

The _class_ of this `dat` object is a `data.frame`. `data.frame`s are R's representation of tabular data. `dat` is also a `tbl_df` ("tibble"), which is the tidyverse's version of a `data.frame` that is a little bit more stringent (i.e. safe) than base R's version, and it has a prettier `print()` method:

```{r}
print(dat)
```

---

# Data types in R

You may have noticed a bunch of three-letter abbreviations next to the variable names in `print(dat)` and the variable content in `str(dat)`. These are the _type signatures_ for each variable:

- Double (`dbl` / `num`): All real numbers, like $50$, $23.193$, or $1.24\times10^{-8}$
- Character (`chr`): Text/strings, written with single- or double-quotes (e.g. `"a string"`)

There are a couple more that you will encounter regularly:

- Integer (`int`): Integers and counting numbers. No decimals and scientific notation here!
- Logical (`lgl`): Boolean data, written as `TRUE` or `FALSE`.
- Factor (`fct`): Factors, R's representation of categorical data. They look like character data with `print()`, but under the hood they are integers. These should only exist in cleaned, ready-for-analysis data!

---

## Quick tip

Often, you don't want to see _all_ rows of a dataset. For quick checks, it's useful to look at the first or last rows. To look at the **first**, use `head()`; to look at the **last**, use `tail()`:

.font12[

```{r}
# First row
head(dat, 1)
```

```{r}
# Last 2 rows
tail(dat, 2)
```

]

---

# Recap

.pull-left[

How do I create a variable `x` that refers to the sentence "Hello, world!" ?

]

.pull-right[

Use 1, 2, 3, or 4 to refer to the code line number.

```{r, eval=FALSE}
x "Hello, world!"
"x" <- Hello, world!
x <- "Hello, world!"
x <- Hello, world!
```

]

---

# Recap

.pull-left[

How do I create a variable `x` that refers to the sentence "Hello, world!" ?

]

.pull-right[

Use 1, 2, 3, or 4 to refer to the code line number.

```{r, eval=FALSE}
x "Hello, world!"
"x" <- Hello, world!
*x <- "Hello, world!"
x <- Hello, world!
```

]

---
class: middle, center

# Working with your data

## ðŸ—„ðŸ”

---

# Basic Data Wrangling

Most data wrangling comes down to **four** basic operations:

- **Filtering**: Looking for certain rows in your dataset
- **Arranging**: Reordering your data with respect to some variables
- **Summarizing**: Aggregating data -- to generate statistics, for example
- **Mutating**: Creating, editing, or deleting entire variables

You can also **group** these operations by variables. For example, finding the average math score in each class would be achieved by _grouping_ by class and then _summarizing_ by taking the mean.

---

# Basic Data Wrangling

In these workshops, **we'll continue to use the Tidyverse**, but I'd be remissed if I didn't show how you'd do these operations in base R as well:

.font12[

| Operation   |                                       Tidyverse |                                    Base R       |
|:------------|------------------------------------------------:|------------------------------------------------:|
| Filtering   | `filter(dat, Cohort == "A")`                    | `dat[dat$Cohort == "A", ]`                      |
| Arranging   | `arrange(dat, Final_grade)`                     | `dat[order(dat$Final_grade), ]`                 |
| Summarizing | `summarize(dat, avg_score = mean(Final_score))` | `data.frame(avg_score = mean(dat$Final_score))` |
| Mutating    | `mutate(dat, failed = Final_score < 60)`        | `dat$failed <- dat$Final_score < 60`            |

]

It's already apparent that the Tidyverse has a much more consistent design than base R, but when we get to grouping, it'll become even more clear to see that base R is pretty confusing to work with for wrangling.

---

# Basic Data Wrangling

All of these Tidyverse functions are provided by the **dplyr** package.

```{r}
library(dplyr)
```

---

## Filtering

Looks for rows with a provided comparison, usually with `<`, `>`, `<=`, `>=`, or `==` (note: two equals!).

```{r}
filter(dat, Cohort == "A")
```

---

## Arranging

Sorts a dataset by a `variable` in _ascending_ order with `arrange(dat, variable)`.

```{r, eval=FALSE}
arrange(dat, Final_score)
```

```{r, echo=FALSE}
head(arrange(dat, Final_score), 6)
```

---

## Arranging

Sorts a dataset in _descending_ order with `arrange(dat, -variable)`.

```{r, eval=FALSE}
arrange(dat, -Final_score)
```

```{r, echo=FALSE}
head(arrange(dat, -Final_score), 6)
```

---

## Summarizing

Calculate some summary function on variables, like `mean()`, `median()`, or `sd()`:

```{r}
summarize(
  dat, 
  avg_score = mean(Final_score), 
  sd_score = sd(Final_score)
)
```

You can generate multiple summaries with multiple `summarize()` parameters.

---

## Mutating

Create, edit, or delete variables in a dataset.

```{r, eval=FALSE}
mutate(
  dat,
  # Deletes the `X1` variable. 
  # NULL is a symbol in R that means "no value" (not "missing"!).
  X1 = NULL,
  # Changes Cohort to an index variable
  Cohort = ifelse(Cohort == "A", 1, 2)
)
```

```{r, echo=FALSE}
head(mutate(
  dat,
  # Deletes the `X1` variable. 
  # NULL is a symbol in R that means "no value" (not "missing"!).
  X1 = NULL,
  # Changes Cohort to an index variable
  Cohort = ifelse(Cohort == "A", 1, 2)
), 2)
```

Note: summarizing can have an arbitrary output variable length. Editing using mutate **must** have the _same length_ as the other columns in the dataset _or length 1_, which is a signal to **dplyr** to set all rows to have the same value).

---

# Combining Operations

These "verbs" (as **dplyr** likes to call them) are **composable** meaning you can chain these commands together, allowing you to create complex data transformations or summaries all in one step.

```{r}
summarize(
  filter(dat, Intervention == 1), 
  avg_final_score = mean(Final_score)
)
```

--

This is ugly. Thankfully, there is a better way of chaining these steps together!

---

.center[

.font12[Introducing...]

]

--

.center[

## The Pipe!

# `%>%`

]

--

The pipe `%>%` is an operator provided by **dplyr**<sup>1</sup> that makes writing nested functions _much_ easier.

.footnote[
[1]: Actually **magrittr**, but that packages has more (esoteric) pipes.
]

---

## The Pipe

Instead of writing

```r
h(g(f(x)))
```

you can write

```r
x %>% 
  f() %>% 
  g() %>% 
  h()
```

--

The pipe works by inserting the result of the left hand side into the **first argument** of the right hand side.

---

# Combining Operations

Revisiting our first chained example,

```{r, eval=FALSE}
summarize(
  filter(dat, Intervention == 1), 
  avg_final_score = mean(Final_score)
)
```

we can now use

```{r}
dat %>% 
  filter(Intervention == 1) %>% 
  summarize(avg_final_score = mean(Final_score))
```

---

## Grouping

Now let's add grouping into the mix, which you can easily do with `group_by(variable)`:

```{r}
dat %>% 
  group_by(Intervention) %>% 
  summarize(avg_final_score = mean(Final_score))
```

---

## Grouping

You can group by multiple variables as well. Just add more variables to `group_by()`:

```{r, eval=FALSE}
dat %>% 
* group_by(Intervention, Cohort) %>% 
  summarize(avg_final_score = mean(Final_score))
```

```{r, echo=FALSE}
dat %>% 
  group_by(Intervention, Cohort) %>% 
  summarize(avg_final_score = mean(Final_score))
```

---

# Recap

.pull-left[

.font12[

How would I find the median baseline stereotype by cohort and gender? For the sake of this question, let's say that `Baseline_stereotype_endorsement` has been renamed to `stereotype`.

]

]

.pull-right[

.font10[

```{r, eval=FALSE}
# 1.
dat %>% 
  filter(Cohort, Female) %>% 
  summarize(
    med_stereotype = median(stereotype)
  )

# 2.
dat %>%
  summarize(
    med_stereotype = median(stereotype)
  ) %>% 
  group_by(Cohort, Female) 

# 3.
dat %>%
  group_by(Cohort, Female) %>% 
  summarize(
    med_stereotype = median(stereotype)
  )

# 4.
dat %>%
  summarize(
    med_stereotype = median(stereotype)
  ) %>% 
  filter(Cohort, Female)
```

]

]

---

# Recap

.pull-left[

.font12[

How would I find the median baseline stereotype by cohort and gender? For the sake of this question, let's say that `Baseline_stereotype_endorsement` has been renamed to `stereotype`.

]

]

.pull-right[

.font10[

```{r, eval=FALSE}
# 1.
dat %>% 
  filter(Cohort, Female) %>% 
  summarize(
    med_stereotype = median(stereotype)
  )

# 2.
dat %>%
  summarize(
    med_stereotype = median(stereotype)
  ) %>% 
  group_by(Cohort, Female) 

# 3.
*dat %>%
* group_by(Cohort, Female) %>% 
* summarize(
*   med_stereotype = median(stereotype)
* )

# 4.
dat %>%
  summarize(
    med_stereotype = median(stereotype)
  ) %>% 
  filter(Cohort, Female)
```

]

]

---

# Revisiting dplyr vs. base R

Just to reinforce the fact that base R does not have a consistent interface, I'd like to show y'all how to do these grouped operations.

--

.center[

### Stand by for a lot of code that doesn't make sense! ðŸ˜³

]

---

## Filtering

```{r, eval=FALSE}
# dat %>% 
#   group_by(Cohort) %>% 
#   filter(Final_score > 85)

partioned <- lapply(
  unique(dat$Cohort), 
  function(cohort) dat[dat$Cohort == cohort, ]
)

filtered <- lapply(
  partioned, 
  function(part) part[part$Final_score > 85, ]
)

do.call(rbind, filtered)
```

---

## Mutating

Grouped mutates can be useful when cleaning longitudinal data per participant, e.g. restricting age to be the baseline observation of the participant.

.font12[

```{r, eval=FALSE}
# dat %>%
#   group_by(Cohort) %>%
#   mutate(median_stereotype = median(Baseline_stereotype_endorsement))

partioned <- lapply(
  unique(dat$Cohort), 
  function(cohort) dat[dat$Cohort == cohort, ]
)

mutated <- lapply(
  partioned, 
  function(part) {
    part$median_stereotype <- median(part$Baseline_stereotype_endorsement)
    part
  }
)

do.call(rbind, mutated)
```

]

---

## Summarizing

.font12[

```{r}
# dat %>% 
#   group_by(Intervention, Cohort) %>% 
#   summarize(avg_final_score = mean(Final_score))

agg <- aggregate(Final_score ~ Intervention + Cohort, data = dat, FUN = mean)
names(agg) <- ifelse(names(agg) == "Final_score", "avg_final_score", names(agg))
agg
```

]

---

## Long story short

Base R _does_ have the capability to do all of these operations out of the box.

--

The problem is you have to know a **lot** of the language to do **anything useful**. This contributes to a steep and unproductive learning curve which dismays all but the most tenacious "leaRners".

--

### I don't want y'all to have that experience!

