---
title: "Session 4"
subtitle: "Workflows in R: Recommended practices"
institute: "Global TIES for Children"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      countIncrementalSlides: false
      highlightLines: true
      highlightStyle: solarized-light
      navigation:
        scroll: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(xaringanthemer)
style_mono_accent(
  text_font_google = google_font("Montserrat", "300", "300i"),
  code_font_google = google_font("Fira Mono")
)
library(magrittr)
library(flair)
```

<style>
:root {
  --font12: 12pt;
  --font10: 10pt;
  --font2: 2pt;
  --nyu-yellow: #ecaa00;
  --nyu-red: #c50f3c;
  --nyu-violet: #57068c;
}

.font12 > table {
  font-size: var(--font12);
}

.font12 > p {
  font-size: var(--font12);
}

.font12 > pre .remark-code-line {
  font-size: var(--font12);
}

.font10 > table {
  font-size: var(--font10);
}

.font10 > p {
  font-size: var(--font10);
}

.font10 > pre .remark-code-line {
  font-size: var(--font10);
}

.font10 > pre {
  font-size: var(--font10);
}

.font2 > pre {
  font-size: var(--font2);
}

.font10 .str_view {
  font-size: var(--font10);
}

.font10.rendered > pre {
  font-size: var(--font10);
  font-family: var(--code-font-family);
  background-color: #fdf6e3;
  color: #657b83;
  padding: 5px;
}

.yellow {
  color: var(--nyu-yellow);
}

.red {
  color: var(--nyu-red);
}

.violet {
  color: var(--nyu-violet);
}

.highlighted {
  background-color: var(--code-highlight-color);
}

.highlighted-block {
  font-family: var(--code-font-family), Menlo, Consolas, Monaco, Liberation Mono, Lucida Console, monospace;
}

</style>

# Last Time

We talked about:

- Pattern matching
- String manipulation
- Function writing
- Iteration

---

# Agenda

Most of the workshop will be more about project structure, however...

By the end of this workshop, you should be able to understand this block of code:

.font10[

```{r, eval=FALSE}
library(drake)

drake_plan(
  raw_data = readxl::read_excel(file_in("raw_data.xlsx")),
  data = raw_data %>%
    mutate(Ozone = replace_na(Ozone, mean(Ozone, na.rm = TRUE))),
  hist = create_plot(data),
  fit = lm(Ozone ~ Wind + Temp, data),
  report = rmarkdown::render(
    knitr_in("report.Rmd"),
    output_file = file_out("report.html"),
    quiet = TRUE
  )
)
```

]

---

# Agenda

We'll be covering

- R-script based project structures
- R "package" project structure
- Controlling code complexity
- **drake**
- Functional programming

I'll have the **tidyverse** meta-package attached throughout the presentation.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

---
class: inverse, middle, center

# Project Structures

## 🗄 🗄 🗄 

---

# Project Structures in R

There are several styles of R project out there, and it can be daunting to choose the "best" one. Like most things in life, **there is no "best" project style**. Each has their pros and cons.

These are the project "styles" we'll cover today:

1. R-script based project
1. R Markdown based project
1. R package project
1. **drake** project

The order of these _roughly_ translates to the size of the project, from the smallest scale to largest.

---

# R-script Projects

The simplest structure is the R script. For very, _very_ small projects, it's sufficient to use a single R script:

.pull-left.font10[

<pre>
project/
  data/
    .. All data are stored here ..
  assets/
    .. All derived assets (reports and whatnot) ..
  make.R
</pre>

]

.pull-right.font10[

```{r, eval=FALSE}
# make.R
library(tidyverse)

dat <- read_csv(here::here("data/dataset.csv"))

# ... Do stuff ...

save_results(results, file = here::here("assets/results.dat"))
```


]

All code is stored in the "make.R" file, and its layout roughly follows:

1. Attach required packages with `library()`
1. Load required datasets
1. Do stuff to the data to get results
1. Save results

Each time you change your code or input data, you `source()` the "make.R" script and rerun the code. As you can imagine, this can get unwieldy pretty quickly...
